---
title: "Assignment13"
author: "James Brock"
date: "26/11/2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(data.table)
```

# Part 1
```{r part1, echo=FALSE}

# a function used to test for two groups whether they are fundamentally different along a given column
# data is the dataset we are studying
# val_col is the variable we wish to investigate
# group_col contains the two groups we are investigating
# var_equal is a boolean value that determines the test we are using
t_test_function <- function(data, val_col, group_col, var_equal){
  # split the data according to the groups, probably could be done more dynamically for general use
  peng_1 <- data[data[[group_col]] == "Adelie",]
  peng_2 <- data[data[[group_col]] == "Chinstrap",]
  
  #get the sample size of each group
  sample_size_1 <- peng_1 %>% nrow()
  sample_size_2 <- peng_2 %>% nrow()
  
  # get the sample mean of each group
  sample_mean_1 <- peng_1 %>% pull(val_col) %>% mean(na.rm=1)
  sample_mean_2 <- peng_2 %>% pull(val_col) %>% mean(na.rm=1)
  
  # calculate the sample sd of each group
  sample_sd_1 <- peng_1 %>% pull(val_col) %>% sd(na.rm = 1)
  sample_sd_2 <- peng_2 %>% pull(val_col) %>% sd(na.rm = 1)
  
  # find the combined standard deviation of the two groups using the previously found sds
  sd_combined <- sqrt(((sample_size_1 - 1)*sample_sd_1^2+(sample_size_2 - 1)*sample_sd_2^2)/(sample_size_1+sample_size_2-2))
  
  # test which test to use now that all common metrics have been calculated
  if(var_equal){
    # perform student unpaired t test
    
    # calculate t statistic
    t_statistic <- (sample_mean_1-sample_mean_2)/(sd_combined*sqrt(1/sample_size_1+1/sample_size_2))
    
    # find the p value
    p_value <- 2*(1-pt(abs(t_statistic), df=sample_size_1+sample_size_2-2))
    
    # generate the effect size, however it seems to not get reported in the final return 
    effect_size <- (sample_mean_1 - sample_mean_2) / sd_combined 
    
    # determine degrees of freedom
    dof <- sample_size_1 + sample_size_2 - 2
    
    # generate a data frame of the calculated values to output
    result_df <- data.frame(t_statistic, dof, p_value)
    colnames(result_df) <- c("t_stat", "dof", "p_val")
    return(result_df)
  } else {
    # perform welch unpaired t test
    
    # generate t statatistic
    t_statistic_welch <- (sample_mean_1 - sample_mean_2) / (sqrt(sample_sd_1^2/sample_size_1 + sample_sd_2^2/sample_size_2))
    
    # compute p value
    p_val_welch <- 2*(1-pt(abs(t_statistic_welch), df = sample_size_1+sample_size_2 - 2))
    
    # find degrees of freedom
    dof_welch <- ((sample_sd_1^2/sample_size_1) + (sample_sd_2^2/sample_size_2))^2 / ((1/(sample_size_1 - 1)* (sample_sd_1^2/sample_size_1)^2) + (1/(sample_size_2 - 1)*(sample_sd_2^2/sample_size_2)^2))
    
    # combine values into a data frame to be returned from the function
    result_df_welch <- data.frame(t_statistic_welch, dof_welch, p_val_welch)
    colnames(result_df_welch) <- c("t_stat", "dof", "p_val")
    return(result_df_welch)
  }
 }

# remove the Gentoo species and remove NA values
peng_AC<-penguins%>%
  drop_na(species,body_mass_g)%>% 
  filter(species!="Gentoo")

# perform t test on our created function on the body mass variable and test it on both tests used
t_test_function(data=peng_AC, val_col="body_mass_g", group_col="species", var_equal = TRUE)
t_test_function(data=peng_AC, val_col="body_mass_g", group_col="species", var_equal = FALSE)

# compare results of our implementation to the built in function
t.test(body_mass_g~species, data=peng_AC, var.equal = TRUE)
t.test(body_mass_g~species, data=peng_AC, var.equal = FALSE)

```

# Part 2
```{r part2, echo = FALSE}
library(Stat2Data)

data("Hawks")
hawks <- data.table(Hawks)# load hawks data

# for a given species of hawk, extract the Hallux column values and remove NA values
hallux_by_species <- function(data, species_code){
  return(data %>% filter(Species == species_code) %>%
    drop_na(Hallux) %>%
    pull(Hallux))
}

# perform wilcox test on the two species CH and SS to determine if there is a difference in their hallux lengths, exact means that an exact p value is calculated
wilcox.test(hallux_by_species(hawks, "CH"), hallux_by_species(hawks, "SS"), exact=TRUE)

# get the sample sizes of each of the two species
sample_size_ss <- length(hallux_by_species(hawks, "SS"))
sample_size_ch <- length(hallux_by_species(hawks, "CH"))

# calculate the U statistic between the two groups based on their hallux lengths
u_statistic <- as.numeric(wilcox.test(hallux_by_species(hawks, "CH"), hallux_by_species(hawks, "SS"), exact=TRUE)$statistic)

# calculate their rank biserial value based on the previously extracted u statistic
rank_biserial <- (2*u_statistic / (sample_size_ss * sample_size_ch)) - 1
rank_biserial

# a function designed to calculate the u statistic and effect size for two groups within a dataset
# based on a specific variable

# df is the data, val_col is the variable of interest
# group_col is the group we are grouping by and group_1 and group_2 are those respective groups
MWW_stats <- function(df, val_col, group_col, group_1, group_2){
  # separate the groups of interest and extract the values in the column of interest
  group_1_vals <- df[df[[group_col]] == group_1, ] %>% pull(val_col) %>% na.omit()
  group_2_vals <- df[df[[group_col]] == group_2, ] %>% pull(val_col) %>% na.omit()
  
  # get their respective sample sizes
  n_1 <- length(group_1_vals)
  n_2 <- length(group_2_vals)
  
  # calculate the u statistic with built in tie handling
  u_stat <- 0
  for (x in group_1_vals){
    for (y in group_2_vals){
      if(x > y){
        u_stat <- u_stat + 1
      } else if (x == y) {
        u_stat <- u_stat + 0.5
      }
    }
  }
  
  # using the calculated u statistic, find the effect size
  effect_size <- 2*u_stat / (n_1*n_2) - 1
  
  # return the results in a dataframe
  result_df_mww <- data.frame(u_stat, effect_size)
  colnames(result_df_mww) <- c("U_stat", "effect_size")
  return(result_df_mww)
}

# call the function to test it works with the penguins data
MWW_stats(penguins, "body_mass_g", "species", "Adelie", "Chinstrap")

```
If we go off the standard significance value of 0.05, assuming the data is i.i.d and roughly gaussian, we can clearly see that given the incredibly small value of p, there is a clear difference between the hallux length of the two species.

The effect size of 0.9466224 along with the rejection of the null hypothesis represents an incredibly significant finding in the difference of hallux sizes between the two species.