---
title: "Assignment13"
author: "James Brock"
date: "26/11/2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(data.table)
```

# Part 1
```{r part1, echo=FALSE}
t_test_function <- function(data, val_col, group_col, var_equal){

  peng_1 <- data[data[[group_col]] == "Adelie",]
  peng_2 <- data[data[[group_col]] == "Chinstrap",]
  
  sample_size_1 <- peng_1 %>% nrow()
  sample_size_2 <- peng_2 %>% nrow()
  
  sample_mean_1 <- peng_1 %>% pull(val_col) %>% mean(na.rm=1)
  sample_mean_2 <- peng_2 %>% pull(val_col) %>% mean(na.rm=1)
  
  sample_sd_1 <- peng_1 %>% pull(val_col) %>% sd(na.rm = 1)
  sample_sd_2 <- peng_2 %>% pull(val_col) %>% sd(na.rm = 1)
  
  sd_combined <- sqrt(((sample_size_1 - 1)*sample_sd_1^2+(sample_size_2 - 1)*sample_sd_2^2)/(sample_size_1+sample_size_2-2))
  
  if(var_equal){
    t_statistic <- (sample_mean_1-sample_mean_2)/(sd_combined*sqrt(1/sample_size_1+1/sample_size_2))
    p_value <- 2*(1-pt(abs(t_statistic), df=sample_size_1+sample_size_2-2))
    
    #unsure where to chuck this in
    effect_size <- (sample_mean_1 - sample_mean_2) / sd_combined 
    
    dof <- sample_size_1 + sample_size_2 - 2
    
    result_df <- data.frame(t_statistic, dof, p_value)
    colnames(result_df) <- c("t_stat", "dof", "p_val")
    return(result_df)
  } else {
    t_statistic_welch <- (sample_mean_1 - sample_mean_2) / (sqrt(sample_sd_1^2/sample_size_1 + sample_sd_2^2/sample_size_2))
    p_val_welch <- 2*(1-pt(abs(t_statistic_welch), df = sample_size_1+sample_size_2 - 2))
    
    # slightly off here
    dof_welch <- ((sample_sd_1^2/sample_size_1) + (sample_sd_2^2/sample_size_2))^2 / ((1/(sample_size_1 - 1)* (sample_sd_1^2/sample_size_1)^2) + (1/(sample_size_2 - 1)*(sample_sd_2^2/sample_size_2)^2))
    
    result_df_welch <- data.frame(t_statistic_welch, dof_welch, p_val_welch)
    colnames(result_df_welch) <- c("t_stat", "dof", "p_val")
    return(result_df_welch)
  }
 }

peng_AC<-penguins%>%
  drop_na(species,body_mass_g)%>%
  filter(species!="Gentoo")

t_test_function(data=peng_AC, val_col="body_mass_g", group_col="species", var_equal = TRUE)
t_test_function(data=peng_AC, val_col="body_mass_g", group_col="species", var_equal = FALSE)

t.test(body_mass_g~species, data=peng_AC, var.equal = TRUE)
t.test(body_mass_g~species, data=peng_AC, var.equal = FALSE)

```

# Part 2
```{r part2, echo = FALSE}
library(Stat2Data)

data("Hawks")
hawks <- data.table(Hawks)

hallux_by_species <- function(data, species_code){
  return(data %>% filter(Species == species_code) %>%
    drop_na(Hallux) %>%
    pull(Hallux))
}

wilcox.test(hallux_by_species(hawks, "CH"), hallux_by_species(hawks, "SS"), exact=TRUE)

sample_size_ss <- length(hallux_by_species(hawks, "SS"))
sample_size_ch <- length(hallux_by_species(hawks, "CH"))
u_statistic <- as.numeric(wilcox.test(hallux_by_species(hawks, "CH"), hallux_by_species(hawks, "SS"), exact=TRUE)$statistic)

rank_biserial <- (2*u_statistic / (sample_size_ss * sample_size_ch)) - 1
rank_biserial

MWW_stats <- function(df, val_col, group_col, group_1, group_2){
  group_1_vals <- df[df[[group_col]] == group_1, ] %>% pull(val_col) %>% na.omit()
  group_2_vals <- df[df[[group_col]] == group_2, ] %>% pull(val_col) %>% na.omit()
  
  n_1 <- length(group_1_vals)
  n_2 <- length(group_2_vals)
  
  u_stat <- 0
  for (x in group_1_vals){
    for (y in group_2_vals){
      if(x > y){
        u_stat <- u_stat + 1
      } else if (x == y) {
        u_stat <- u_stat + 0.5
      }
    }
  }
  
  effect_size <- 2*u_stat / (n_1*n_2) - 1
  
  result_df_mww <- data.frame(u_stat, effect_size)
  colnames(result_df_mww) <- c("U_stat", "effect_size")
  return(result_df_mww)
}

MWW_stats(penguins, "body_mass_g", "species", "Adelie", "Chinstrap")

```
If we go off the standard significance value of 0.05, assuming the data is i.i.d and roughly gaussian, we can clearly see that given the incredibly small value of p, there is a clear difference between the hallux length of the two species.

The effect size of 0.9466224 along with the rejection of the null hypothesis represents an incredibly significant finding in the difference of hallux sizes between the two species.